---
title: "Lab 5"
author: "Bernard Fay, Github: bj99fay; Chase Weida, GitHub: cweida"
date: "4/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Load the data
```{r}
accident <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
person <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```

## 1
```{r}
day.dist <- accident %>%
  mutate(DAY_WEEK_NAME=recode(as.factor(DAY_WEEK), "1" = "SUNDAY", "2" = "MONDAY", "3" = "TUESDAY", "4" = "WEDNESDAY", 
         "5" = "THURSDAY", "6" = "FRIDAY", "7" = "SATURDAY", "9" = "UNKNOWN")) %>%
  group_by(DAY_WEEK_NAME)
day.dist %>% ggplot(aes(x=DAY_WEEK_NAME)) + geom_bar() + labs(x = "DAy of Week", title = "Distribution of Accidents by Day of Week")
```

Accidents happen more often on Fridays, Saturdays, and Sundays. These are days that people don't typically work, so this could be a cause of the higher number of accidents.

## 2
```{r}
time.dist <- accident %>%
  filter(HOUR < 24)
  group_by(HOUR)
time.dist %>% ggplot(aes(x=HOUR)) + geom_bar() + labs(x = "Time of Day by Hour (24hr clock)", title = "Distribution of Accidents by Time of Day")
```

Accidents tend to happen in the late afternoon/evening. This is the time most people head home from work, so this could be a potential cause of the increase i accidents.

## 3
```{r}
drunk <- accident %>%
  mutate(DRUNK = ifelse(DRUNK_DR > 0, "Yes", "No"))
  group_by(DRUNK)

drunk %>% filter(DRUNK == "Yes") %>% count() %>% rename("Drunk driving accidents" = "n")
  
drunk %>% ggplot(aes(x=DRUNK)) + geom_bar() + labs(x = "Was a Drunk Driver Invloved?", title = "Accidents Invloving >= 1 Drunk Driver")
```

There were 8769 accidents involving at least one drunk driver. There are around three times as many accidents not involving a drunk driver, so drunk driving account fro around 1/4 of all accidents.

## 4
```{r}
person <- person %>%
  filter(PER_TYP == 1)
person
```

